Evolution of precopulatory and postcopulatory inbreeding behaviour

### 2016-09-06 ms_pre_post.Rmd
Revise Model analysis to add new 'hypotheses'

<!--- TODO: PARAGRAGH ON EVOLUTION OF POLYANDRY --->
Critically, evolution of post-copulatory inbreeding strategy is precluded if females only mate with one male, meaning that evolution of polyandry is intrinsically tied to female inbreeding fertilisation strategy. Although polyandry is commonplace in wild populations, the regularity with which it is observed in the wild remains theoretically puzzling because engaging in polyandry does not necessarily increase the number of offspring that a female produces, meaning that the fitness benefits associated with polyandry are not always clear [@Parker2013; @Taylor2014]. It has been hypothesised that females can acheive a net fitness benefit if polyandry provides them with additional opportunities to adjust their inbreeding through copulations with additional males [e.g., @Tregenza2002; @Akcay2007; @Lehtonen2015]. @Duthie showed that selection for polyandry to facilitate inbreeding avoidance is sometimes predicted, but only under highly constrained conditions in which very few males are available during a female's initial mate choice but many males are available during a female's additional mate choice(s), or when expression of polyandry is conditionally expressed based on the degree of inbreeding in a female's initially chosen mate. However, @Duthie assumed that polyandrous females had no control over the assignment of paternity to their offspring. If after engaging in polyandry, females could bias paternity toward non-relatives, then selection for polyandry to avoid inbreeding might be more widely predicted [@Zeh1997]. However, no models quantify the degree to which expression of biased paternity resulting from evolving inbreeding fertilisation strategy can feedback to drive evolution of polyandry in populations of potentially inbreeding females and males with biologically realistic distributions of relatedness.

@@

<!--- TODO: PARAGRAGH ON ORDER OF INTERACTIONS --->
Selection for inbreeding mating strategy versus inbreeding fertilisation strategy and associated polyandry might also depend evolutionary constraints, or on existing phentypes in the population. For example, if phenotypes biasing fertilisation to avoid inbreeding are precluded by evolutionary constraints [@Bradshaw1991; @Futuyma2010], then inbreeding avoidance might only be possible before copulation occurs, perhaps strengthening selection for inbreeding mating strategy and weakening selection for polyandry. Alternatively, evolution of female inbreeding mating strategy might be precluded due to severely limited kin recognition or the inability to resist male coercion given sexual conflict [@Arnqvist2005a; @Parker2006; @Penn2010]. In such conditions, females might only be able to express choice through inbreeding fertilisation strategy, perhaps strengthening selection for inbreeding avoidance through polyandry and post-copulatory mechanisms. Selection for inbreeding mating or fertilisation strategy might also be weak or negligible if the strategy is redundant due to the presence of an already established adaptive inbreeding strategy (e.g., if females already successfully avoid inbreeding during mate encounters, then they might be unable to increase their fitness further by biasing fertilisation after mating has occurred). It is therefore necessary to consider evolution of inbreeding mating strategy and inbreeding fertilisation strategy when each evolves in a population where its complementary inbreeding strategy is evolutionarily constrained or already fixed as an adaptation.

<!--- TODO: OVERVIEW OF THE POINT OF THE MODELLING --->
Here we apply a model of female and male reproductive interactions and associated inbreeding depression in offspring viability to quantify evolution of three separate traits including female tendency for polyandry, female pre-copulatory inbreeding strategy (mating strategy), and female post-copulatory inbreeding strategy (fertilisation strategy). We use individual-based modelling and *in silico* experiments to test the hypotheses that, given inbreeding depression in offspring viability, evolution of post-copulatory inbreeding avoidance causes selection for evolution of polyandry, and that pre-copulatory but not post-copulatory inbreeding avoidance will evolve when pre-copulatory inbreeding avoidance is less costly than the combined cost of post-copulatory inbreeding avoidance and polyandry, while post-copulatory but not pre-copulatory inbreeding avoidance will evolve when pre-copulatory inbreeding avoidance is more costly than the combined costs of post-copulatory inbreeding avoidance and polyandry. We thereby provide a framework for predicting evolution of pre-copulatory and post-copulatory inbreeding avoidance across reproductive systems with different relative costs and different magnitudes of inbreeding depression in offspring viability.

@@

To analyse our model, we track mean values of alleles underlying tendency for polyandry ($P_{a}$), inbreeding mating strategy ($M_{a}$), and inbreeding fertilisation strategy ($F_{a}$) in a focal population. We present evolutionary dynamics of these mean allele values across different combinations of inbreeding depression slopes and costs to test our two focal hypotheses. The combinations of costs and inbreeding depression slopes that we consider include all eight possible permutations of values of costs of tendency for polyandry ($c_{P}$), inbreeding mating strategy ($c_{M}$), and inbreeding fertilisation strategy ($c_{F}$) equal to either $0$ or $0.02$ [@Duthie observed $0.02$ to be strong cost of tendency for polyandry], and for values of $\beta = 3$ ($8$ total combinations). In all of these simulations, we record mean values of $P_{a}$, $M_{a}$, $F_{a}$ in each generation. Each of the 16 possible cost combinations is replicated $40$ times, and grand mean values and standard deviation of means is calculated in each generation across replicate simulations. These analyses allowed us to infer how allele values change over generations in response to costs, but also in response to the changing values of other alleles and therefore potentially interacting phenotypes (e.g., selection on inbreeding fertilisation strategy might be weak if polyandry is rare).

@@

To test hypotheses regarding evolution under different relative costs of phenotypes, we consider all combinations of costs and inbreeding depression parameter values using only simulations in which all three traits evolve. To test our hypothesis that inbreeding mating strategy will evolve toward inbreeding avoidance ($M_{p}<0$) when mating costs are low relative to the costs of tendency for polyandry and inbreeding fertilisation strategy, but that selection for inbreeding fertilisation strategy will then be negligible ($F_{p} \approx 0$), we quantify evolution of $M_{p}$ and $F_{p}$ over generations in simulations where $c_{M}=0$ but $c_{P}$, $c_{F}$, or both equals $0.02$. We compare evolution in these three cost parameter value combinations with simulations in which all costs equal either $0$ or $0.02$. If our hypothesis is supported, then $M_{p}$ but not $F_{p}$ will evolve to more negative values when $c_{M}$ is relatively low than when all cost values are uniform.

To test the complementary hypothesis that inbreeding fertilisation strategy will evolve toward inbreeding avoidance ($F_{p}<0$) given low costs of polyandry and inbreeding fertilisation strategy relative to the cost of inbreeding mating strategy, but that selection on inbreeding mating strategy will be negligible ($M_{p} \approx 0$), we qualtify evolution of $M_{p}$ and $F_{p}$ over generations when $c_{F}=0$, $c_{P}=0$, and $c_{M}=0.02$. We compare evolution for this combination of cost parameter values with simulations in which all costs equal either $0$ or $0.02$. If our hypothesis is supported, then $F_{p}$ but not $M_{p}$ will evolve to more negative values when $c_{F}$ is relative low than when all cost values are uniform.

>>>>

<!--- TODO: PARAGRAGH ON EVOLUTION OF POLYANDRY (H1) --->
Critically, evolution of post-copulatory inbreeding strategy is precluded if females only mate with one male, meaning that evolution of polyandry is intrinsically tied to female inbreeding fertilisation strategy. Although polyandry is commonplace in wild populations, the regularity with which it is observed in the wild remains theoretically puzzling because engaging in polyandry does not necessarily increase the number of offspring that a female produces, meaning that the fitness benefits associated with polyandry are not always clear [@Parker2013; @Taylor2014]. It has been hypothesised that females can acheive a net fitness benefit if polyandry provides them with additional opportunities to adjust their inbreeding through copulations with additional males [e.g., @Tregenza2002; @Akcay2007; @Lehtonen2015]. @Duthie showed that selection for polyandry to facilitate inbreeding avoidance is sometimes predicted, but only under highly constrained conditions in which very few males are available during a female's initial mate choice but many males are available during a female's additional mate choice(s), or when expression of polyandry is conditionally expressed based on the degree of inbreeding in a female's initially chosen mate. However, @Duthie assumed that polyandrous females had no control over the assignment of paternity to their offspring. If after engaging in polyandry, females could bias paternity toward non-relatives, then selection for polyandry to avoid inbreeding might be more widely predicted [@Zeh1997]. However, no models quantify the degree to which expression of biased paternity resulting from evolving inbreeding fertilisation strategy can feedback to drive evolution of polyandry in populations of potentially inbreeding females and males with biologically realistic distributions of relatedness.

@@

<!--- TODO: PARAGRAGH ON ORDER OF INTERACTIONS --->
Selection for inbreeding mating strategy versus inbreeding fertilisation strategy and associated polyandry might also depend evolutionary constraints, or on existing phenotypes in the population. For example, if phenotypes biasing fertilisation to avoid inbreeding are precluded by evolutionary constraints [@Bradshaw1991; @Futuyma2010], then inbreeding avoidance might only be possible before copulation occurs, perhaps strengthening selection for inbreeding mating strategy and weakening selection for polyandry. Alternatively, evolution of female inbreeding mating strategy might be precluded due to severely limited kin recognition or the inability to resist male coercion given sexual conflict [@Arnqvist2005a; @Parker2006; @Penn2010]. In such conditions, females might only be able to express choice through inbreeding fertilisation strategy, perhaps strengthening selection for inbreeding avoidance through polyandry and post-copulatory mechanisms. Selection for inbreeding mating or fertilisation strategy might also be weak or negligible if the strategy is redundant due to the presence of an already established adaptive inbreeding strategy (e.g., if females already successfully avoid inbreeding during mate encounters, then they might be unable to increase their fitness further by biasing fertilisation after mating has occurred). It is therefore necessary to consider evolution of inbreeding mating strategy and inbreeding fertilisation strategy when each evolves in a population where its complementary inbreeding strategy is evolutionarily constrained or already fixed as an adaptation.

<!--- TODO: OVERVIEW OF THE POINT OF THE MODELLING --->
Here we apply a model of female and male reproductive interactions and associated inbreeding depression in offspring viability to quantify evolution of three separate traits including female tendency for polyandry, female pre-copulatory inbreeding strategy (mating strategy), and female post-copulatory inbreeding strategy (fertilisation strategy). We use individual-based modelling and *in silico* experiments to test the hypotheses that, given inbreeding depression in offspring viability, evolution of post-copulatory inbreeding avoidance causes selection for evolution of polyandry. Additionally, we test the hypothesis that costs associated with polyandry and inbreeding strategies will strongly affect evolution of each such that that pre-copulatory but not post-copulatory inbreeding avoidance will evolve when pre-copulatory inbreeding avoidance is less costly than the combined cost of post-copulatory inbreeding avoidance and polyandry, while post-copulatory but not pre-copulatory inbreeding avoidance will evolve when pre-copulatory inbreeding avoidance is more costly than the combined costs of post-copulatory inbreeding avoidance and polyandry. Finally, we test the hypothesis that constraints on evolution of one inbreeding strategy (e.g., inbreeding mating strategy) will increase the strength of selection on evolution of a complementary inbreeding strategy (e.g., inbreeding fertilisation strategy), and that evolution of one inbreeding strategy will be precluded if it is introduced into a population that is already fixed for a complementary inbreeding strategy that is adaptive. We thereby provide a framework for predicting evolution of pre-copulatory and post-copulatory inbreeding avoidance across reproductive systems with different relative costs and different magnitudes of inbreeding depression in offspring viability.

@@

To analyse our model, we track mean values of alleles underlying tendency for polyandry ($P_{a}$), inbreeding mating strategy ($M_{a}$), and inbreeding fertilisation strategy ($F_{a}$) in a focal population. We present evolutionary dynamics of these mean allele values across different combinations of inbreeding depression slopes and costs to test our four focal hypotheses (see below). The combinations of costs and inbreeding depression slopes that we consider include all eight possible permutations of values of costs of tendency for polyandry ($c_{P}$), inbreeding mating strategy ($c_{M}$), and inbreeding fertilisation strategy ($c_{F}$) equal to either $0$ or $0.02$ [@Duthie observed $0.02$ to be strong cost of tendency for polyandry], and for values of $\beta = 3$ ($8$ total combinations). In all of these simulations, we record mean values of $P_{a}$, $M_{a}$, $F_{a}$ in each generation. Each of the $8$ possible cost combinations is replicated $40$ times, and grand mean values and standard deviation of means is calculated in each generation across replicate simulations. These analyses allowed us to infer how allele values change over generations in response to costs, but also in response to the changing values of other alleles and therefore potentially interacting phenotypes (e.g., selection on inbreeding fertilisation strategy might be weak if polyandry is rare).

@@

To test hypotheses regarding evolution under different relative costs of phenotypes, we consider all combinations of costs and inbreeding depression parameter values using only simulations in which all three traits evolve. To test our hypothesis that inbreeding mating strategy will evolve toward inbreeding avoidance ($M_{p}<0$) when mating costs are low relative to the costs of tendency for polyandry and inbreeding fertilisation strategy, but that selection for inbreeding fertilisation strategy will then be negligible ($F_{p} \approx 0$), we quantify evolution of $M_{p}$ and $F_{p}$ over generations in simulations where $c_{M}=0$ but $c_{P}$, $c_{F}$, or both equals $0.02$ and $\beta=3$. We compare evolution in these three cost parameter value combinations with simulations in which all costs equal either $0$ or $0.02$. If our hypothesis is supported, then $M_{p}$ but not $F_{p}$ will evolve to more negative values when $c_{M}$ is relatively low than when all cost values are uniform. To test the hypothesis that inbreeding fertilisation strategy will evolve toward inbreeding avoidance ($F_{p}<0$) given low costs of polyandry and inbreeding fertilisation strategy relative to the cost of inbreeding mating strategy, but that selection on inbreeding mating strategy will be negligible ($M_{p} \approx 0$), we qualtify evolution of $M_{p}$ and $F_{p}$ over generations when $c_{F}=0$, $c_{P}=0$, and $c_{M}=0.02$ and $\beta=3$. We compare evolution for this combination of cost parameter values with simulations in which all costs equal either $0$ or $0.02$. If our hypothesis is supported, then $F_{p}$ but not $M_{p}$ will evolve to more negative values when $c_{F}$ is relatively low than when all cost values are uniform.

To test the hypothesis that constraints on evolution of one inbreeding strategy will increase the strength of selection on a complementary strategy, we fixed inbreeding strategy phenotypes to values of zero, thereby causing either random mating ($M_{p}=0$) or the random fertilisation of offspring ($F_{p}=0$) according to a fair raffle among mates with respect to kinship. We then compare evolution of the non-fixed trait to its evolution in simulations where all traits are allowed to evolve. In all simulations testing this hypothesis, we used strong inbreeding depression ($\beta=3$) and included $40$ replicates for each set of parameter values.

To test the hypothesis that the existence of an adaptive inbreeding strategy will decrease the strength of selection on, and thereby preclude the invasion of, a complementary strategy, we first used exploratory simulations to determine evolution of inbreeding mating strategy, and inbreeding fertilisation strategy and associated polyandry, in isolation. To test whether or not inbreeding mating avoidance was predicted to evolve when adaptive polyandry to avoid inbreeding was already established, we initiated inbreeding mating strategy allele values at zero ($M_{a}=0$), but fixed inbreeding fertilisation strategy allele values at $F_{a}=-5$ and polyandry allele values at $P_{a}=1$. Similarly, to test whether or not inbreeding fertilisation strategy evolved to avoid inbreeding, we initiated inbreeding fertilisation strategy allele values and polyandry allele values at zero ($F_{a}=0$; $P_{a}=0$), but fixed inbreeding mating strategy allele values at $M_{a}=-5$. For all simulations testing this hypothesis, we used strong inbreeding depression ($\beta=3$) and included $40$ replicates for each set of parameter values
